Thesis Title

### 2022-03-07 chapters/introduction.tex
minor fixes

The Global Burden of Disease (GBD) 2019 study reported that microbial pathogens are responsible for more than 400 million years of life lost annually across the globe, a higher burden than either cancer or cardiovascular disease \citep{vos_global_2020}. In particular, lower respiratory infections, diarrhoeal diseases, HIV/AIDS and tuberculosis were amongst the five leading causes of global total years of life lost. More recently, the COVID-19 pandemic, declared as such by the World Health Organization (WHO) on 11 March 2020 after the emergence and global spread of the severe acute respiratory syndrome coronavirus 2 (SARS-CoV-2), and as of January 2022 has caused more than 5.63 million deaths worldwide \citep{ritchie_coronavirus_2020}, making it one of the deadliest pandemics in history. Coronavirus has been responsible for three of the eighteen major pandemics registered throughout modern history \citep{piret_pandemics_2021}, all occurring after the year 2000. \textit{Yersinia pestis}, responsible for three pandemics of plague, \textit{Vibrio cholerae}, with seven cholera pandemics, and Influenza A virus, the causative agent of five flu pandemics, are responsible for the remaining ones, with Influenza being the only other pathogen with a pandemic registered after 2000. Recent decades have also witnessed the emergence of additional virulent pathogens, including the Ebola virus, West Nile virus, Dengue virus and Zika virus, particularly in lower-income countries.

@@

The past few decades have seen a major revolution in the operation of microbial laboratories, driven by the development of molecular technologies and ways to make these accessible, namely amplification-based polymerase chain reaction (PCR), matrix-assisted laser desorption/ionisation - time of flight (MALDI-TOF) and DNA-microarray-based hybridisation technology. These are used in conjunction with traditional techniques such as microscopy, culture and serology, not fully replacing them. Application of these methods differs by suspected infection type: bacterial, viral, fungal or parasitic. for the purpose of this dissertation work, we will be focusing on bacterial and viral infections.

@@

The current gold standard for bacterial pathogen identification in diagnostic microbiology laboratories involves the isolation of the pathogen through culture followed by biochemical testing, a multi-step process that can take days to weeks before obtaining results, depending on the fastidiousness of the organism and if it can be cultured \citep{muhamad_rizal_advantages_2020, giuliano_guide_2019, muhamad_rizal_advantages_2020}. Although culture allows the identification of a wide variety of organisms, some pathogens can escape routine investigation due to strict metabolic necessities for growth or the requirement for specific biochemical tests needed for their identification. Additionally, results will be obscured if a mixed culture is obtained, particularly if the cultures are obtained from sites with a microbiota, such as the gut and the skin, increasing the risk of contamination from normal flora, leading to false results \citep{giuliano_guide_2019}. After successful growth in culture, Gram staining and MALDI-TOF mass spectrometry are often used for identification with good accuracy as long as the pathogen is presented in the the coexisting database \citep{patel_maldi-tof_2015}. An alternate rapid identification method is PCR where nucleic acid fragments are detected through specific probes, being highly sensitive and specific, to the point where PCR may detect bacteria that are not viable after a patient has been treated for an infection \citep{scerbo_beyond_2016}. It's limited to the probes used, with multiplex PCR, an extension of PCR by using multiple primers, tried to address this issue by allowing for the identification of multiple bacteria and other important information such as the detection of antibiotic resistance genes or virulence genes \citep{giuliano_guide_2019}

@@

Choosing an appropriate bacterial typing technique for epidemiological studies depends on the resources available and the minimum intended resolution, ranging from DNA fingerprinting to multilocus sequence typing, Pulsed-field gel electrophoresis and Sequence-based typing \citep{allerberger_molecular_2012,foxman_choosing_2005}. DNA macrorestriction analysis by pulsed-field gel electrophoresis (PFGE), which revolutionised precise separation of DNA fragments, became the most widely implemented DNA fingerprinting technique \citep{allerberger_molecular_2012}, becoming the golden standard for bacterial typing \citep{neoh_pulsed-field_2019}.

In the early 2000s, Multilocus sequence typing (MLST) was proposed as a portable, universal, and definitive method for characterising bacteria \citep{maiden_multilocus_2006}. Instead of enzyme restriction of bacteria DNA, separation of the restricted DNA bands using a pulsed-field electrophoresis chamber, followed by clonal assignment of bacteria based on PFGE banding patterns, MLST relies on the amplification through PCR sequences of internal fragments of housekeeping genes (usually 5 to 7), approximately 450-500 basepairs (bp) in size, followed by its the sequence, usually my Sanger methods. For each house-keeping gene, the different sequences present within a bacterial species are assigned as distinct alleles and, for each isolate, the alleles at each of the (usually) seven loci define the allelic profile or sequence type \citep{larsen_multilocus_2012}. As with PFGE, different schemes, defining what house-keeping gene fragments are used, are available depending on the species. Unlike PFGE, the provision of freely accessible, curated databases of MLST nucleotide sequence data enables the direct comparison of bacterial isolates, providing the basis of a common language for bacterial typing \citep{maiden_multilocus_2006}. So far, MLST schemes for 115 bacterial organisms have been published and made freely available (\url{https://pubmlst.org/organisms}, \cite{jolley_open-access_2018})

@@

Direct detection methods negate the need for virus propagation, detecting the virus directly from the suspect source through nucleic acid and immunological methods. PCR and reverse transcription-PCR are widely applied methods for the detection of both DNA and RNA viruses, respectively, driven by increased awareness of the clinical value of, and demand for, prompt information about viral loads, viral sequence data, and potential antiviral resistance information \citep{cassedy_virus_2021}. Syndromic testing, utilising multiplex PCR to simultaneously amplify the nucleic acid of multiple targets in a single reaction, is now fully integrated into the standard testing practices of many clinical laboratories \citep{dien_bard_panels_2020}. Limitations of these assays include no detection of off-target pathogens, a lack of full susceptibility information, cost, and false-positive results.

Real-time quantitative PCR (qPCR) remains the front line tool in aetiological diagnosis, measuring the production of the target amplicon throughout the reaction and providing quantitative results with high specificity and sensibility, albeit with a significant cost due to sophisticated apparatus despite high-throughput systems being widely established \citep{cassedy_virus_2021}. Immunoassays employ singular-epitope specificity antibodies as the primary means to detect viruses within a sample and provide a much more cost-efficient alternative to nucleic acid detection \citep{cassedy_virus_2021}. One major application is seroprevalence assays, an essential technique for identifying patients who have been exposed to a virus (historical exposure), detecting asymptomatic infection or evaluating vaccine efficacy  \citep{chan_determining_2021, bobrovitz_global_2021}. Lateral flow immunoassays (LFIA) are extensively used for detecting virus-associated protein directly from the source through labelled antibodies binding to their cognate antigens, usually read by way of a colour change at a test line. Besides being very cost-effective, LFAIs have a turnaround time of minutes and the colour change can be observed with the naked eye, therefore facilitating rapid diagnosis but its results are limited to semi-quantitative and it does not typically achieve sensitivity comparable to nucleic-acid detection \citep{estrela_lateral_2016, cassedy_virus_2021, di_nardo_ten_2021}.

@@

Since the publication of the first complete microbial genome a quarter of a century ago, that of the bacterium \textit{Haemophilus influenzae} \citep{hood_dna_1996}, genomics has transformed the field of microbiology, and in particular its clinical application. The paper describing the DNA-sequencing method with chain-terminating inhibitors used in the sequencing of the first microbial genome \citep{sanger_dna_1977}, which earned the late Frederick Sanger his share of the 1980 Nobel Prize in Chemistry alongside Walter Gilbert, was, in 2014, the top fourth in the number of citations with 60335, highlighting its impact in the field of biological sciences, and by extension medicine \citep{van_noorden_top_2014}. Currently, this number has increased to 84546 according to PubMed Central\textsuperscript{\small\textregistered} (PMC) (\url{https://pubmed.ncbi.nlm.nih.gov/, https://www.ncbi.nlm.nih.gov/pmc/articles/PMC431765/}). Since its emergence, reductions in cost, technical advances in sequencing technologies and new computational developments have made genomic sequencing one of the most influential tools in biomedical research, yielding unprecedented insights into microbial evolution and diversity, and the complexity of the genetic variation in both commensal and pathogenic microbes.

The emerging application of genomic technologies in the clinic to combat infectious diseases is transforming clinical diagnostics and the detection and surveillance of outbreaks.

@@

The latter is defined by their use of terminator molecules that are similar to those used in the first-generation of sequencing, preventing elongation of the DNA molecule, but unlike the first methods, it's reversible. To begin the process, a DNA template is primed by a sequence that is complementary to an adapter region, which will initiate polymerase binding to this double-stranded DNA region. During each cycle, a mixture of all four individually labelled and 3'-blocked dNTPs are added. After the incorporation of a single dNTP to each elongating complementary strand, unbound dNTPs are removed and the surface is imaged to identify which dNTP was incorporated at each cluster by optical capture. The fluorophore and blocking group can then be removed and a new cycle can begin \citep{goodwin_coming_2016}. The Illumina systems, which use this technology, accounts for the largest market share for sequencing instruments compared to other platforms\footnote{\url{https://www.forbes.com/companies/illumina/?sh=774358a91aa6}}, allowing paired-end sequencing and having the highest throughput (from 25 million reads for a MiSeq instrument to  1.2 billion reads for a NextSeq instrument\footnote{\url{https://www.illumina.com/systems/sequencing-platforms.html}}), with read lengths ranging from 45 to 300 bases in length with high accuracy, albeit with long running times (4 to 55 hours), rendering this technology a good choice for many sequencing applications where large read length is not required \citep{loman_performance_2012, gupta_chapter_2014, hagemann_overview_2015}.

@@

Despite their wide adoption, second-generation methods require library preparation and an enrichment or amplification step. These steps are time-consuming, introduce biases related to preferential capture or amplification of certain regions, and produce reads with relatively small size, making transversing repetitive genomic regions impossible if they are larger than the read length \citep{hagemann_overview_2015}. Third-generation sequencing technologies, also known as long-read sequencing or single-molecule sequencing, are characterised by the generation of ultra-long-reads, albeit at a much lower throughput than the second-generation \citep{hoang_long-reads-based_2022}. They also have the potential to go beyond four-base sequencing to reveal genome-wide patterns of methylation and other chemical modifications that control the biology of bacteria or the virulence of pathogens \citep{korlach_going_2012}. Currently, commercial long-read sequencing is supported by two companies: Pacific Biosciences and Oxford Nanopore Technologies.

@@

Whole-genome sequencing (WGS) is becoming one of the most widely used applications of microbial genome sequencing. The major advantage of WGS is to yield all the available DNA information content on isolates in a single rapid step following culture (sequencing without culture will be discussed in the \secref{ssec:metagenomics}). In principle, after obtaining a pure culture, either bacterial (see \secref{sssec:bacterial}), viral (see \secref{sssec:viral}) of fungal (see \secref{sssec:fungal}), the data from sequencing contain all the information currently used for diagnostic and typing needs, and much more, thus opening the prospect for large-scale research into pathogen genotype-phenotype associations from routinely collected data \citep{didelot_transforming_2012}.
The cost of producing massive amounts of information requires a new framework with expert handling and processing of computer-driven genomic information, as well as capable computational infrastructures, but through this technology, researchers and clinicians can obtain the most comprehensive view of genomic information and associated biological implications, transforming clinical diagnostics and the detection and surveillance of outbreaks. \citep{cirulli_uncovering_2010, noauthor_genomic_2019, goodwin_coming_2016}.

Targeted sequencing is also proving invaluable to clinical microbial and research, not only by allowing more individual samples to be sequenced within a single run, significantly reducing costs and the amount of data generated while, but also, due to the smaller target size, obtaining results with very high confidence due to the high coverage obtained \citep{goodwin_coming_2016}. This has been particularly useful in viral genomics where sections, such as the capsid, or the complete viral genome can be selectively targeted directly from the suspected sample, offering a more time-effective method to achieve the same output as traditional nucleic acid amplification methods \citep{cassedy_virus_2021}.

@@

WGS has been used in the routine laboratory workflow when typing of pathogens by a method having the highest possible discriminatory power is required either through single nucleotide polymorphism (SNP) or core-genome/whole genome MLST analysis, for example during hospital outbreaks \citep{tagini_bacterial_2017}. Additionally, in bacterial diagnostics, WGS can be used to reveal the presence of AMR genes, or genes associated with virulence and pathogenicity, as well as to discover new genetic mechanisms for the three previously defined important clinical features of a bacterium \citep{rossen_practical_2018}. The implementation of WGS in routine diagnostics requires several adaptations in the laboratory workflow, from the ‘wet’ laboratory part (extraction, library preparation, sequencing), to the ‘dry’ bioinformatics part where genomic data is analysed and its results interpreted by specialised personnel \citep{rossen_practical_2018}.

@@

In outbreak detection and surveillance, genetic sequencing techniques combined with epidemiological data have undoubtedly provided immeasurable insights regarding evolutionary relationships and transmission pathways in various environments \citep{beckett_pandemic_2021, lancet_genomic_2021}. In a pandemic setting, this approach, although not novel, has been revolutionary, particularly in the COVID-19 setting.

In the 2009 swine-origin Influenza A H1N1 pandemic, the first complete genome was publicly available on the 25 of April of 2009 (GenBank accession number FJ966079), about a month after records of increased flu activity in Mexico and 10 days after the first confirmed cases in California, United States of America \citep{smith_origins_2009, novel_swine-origin_influenza_a_h1n1_virus_investigation_team_emergence_2009}. By the time the pandemic was declared, on 11 of June of 2009, \cite{smith_origins_2009} reported the origins and evolutionary genomics of the pandemic influenza A variant with a collection of 813 complete influenza genome sets, 17 of which belonging to the newly swine influenza viruses (GenBank accessions numbers GQ229259–GQ229378). The MERS pandemic, declared as such in 2015 \citep{piret_pandemics_2021}, had its first publicly available sequence on 5 of July 2015 (GenBank accession number KT006149)\citep{lu_complete_2015}, with a sequence from a camel, thought to be an intermediate host for the virus, available as early as 7 of March 2016 (GenBank accession number KU740200) \citep{kandeil_complete_2016, al-shomrani_genomic_2020}.

@@

Despite the increasing adoption of DNA sequencing methods in clinical microbiology, the sequencing of genetic material from a pure culture requires a priori knowledge of what to expect from a particular clinical sample or patient \citep{schuele_future_2021}. In most cases, a priori knowledge is enough to request the most appropriate test, such as multiplexed panels or specific culture media, but this is not always the case. In recent years, there has been a growing interest in using metagenomics to deliver culture-independent approaches to microbial ecology, surveillance and diagnosis \citep{loman_twenty_2015, loman_high-throughput_2012}.
Metagenomic DNA sequence allows detailed characterisation of pathogens in all kinds of samples originating from humans, animals, food and the environment, ligating the diagnostics to surveillance in a true "one health" fashion \citep{rossen__2018}. Unlike PCR or microarrays, it usually does not require primer or probe design, it can be easily multiplexed, and the specificity and selectivity of the sequencing can be adjusted computationally after acquiring the data \citep{dunne_next-generation_2012}.  While most molecular assays target only a limited number of pathogens, metagenomic approaches characterise all DNA or RNA present in a sample, enabling analysis of the entire microbiome as well as the human host genome or transcriptome in patient samples \citep{chiu_clinical_2019}. Whether or not it can entirely replace routine microbiology depends on several conditions and future developments, both technological and computational (see \secref{sec:bioinformatics}).

@@

Quality assessment and control is a basal step to any analysis, and aims to remove and/or filter low quality and low complexity reads, trim adapters, and remove host sequences from the samples’ raw data. There are many tools available but the most commonly used are FastQC\footnote{\url{https://www.bioinformatics.babraham.ac.uk/projects/fastqc/}} (Babraham Bioinformatics) for quality control, followed by Trimmomatic \citep{bolger_trimmomatic_2014}, Cutadapt \citep{martin_cutadapt_2011} or fastp \citep{chen_fastp_2018} to trim and/or filter adaptors, low quality and low complexity sequences. For long-read sequencing, tools like NanoPlot and NanoStats \citep{de_coster_nanopack_2018}, and Filtlong\footnote{\url{https://github.com/rrwick/Filtlong/}} can perform basic quality assessment and control, adapter trimming and low quality trimming respectively.

@@

Metataxonomics (see \secref{sssec:metataxonomics}) is the most widely used technique for microbial diversity analysis \citep{hilton_metataxonomic_2016}, and due to its particularities, the analysis of this data is also very particular.

The same approach of direct taxonomic classification can be followed as with genomic and shotgun metagenomic data, given that the databases include rRNA sequences, but more traditional approaches rely on the identification of OTUs. This is done through sequence matching with the appropriate databases  (see \secref{sssec:metataxonomics}), followed by statistical analysis to assess differences between given groups. Quiime2\footnote{\url{https://qiime2.org/}} \citep{bolyen_reproducible_2019} has become the \textit{de facto} tool for metataxonomic analysis as a framework with an ever-growing suite of plugins and intuitive data visualisation tools for the assessment of results.

>>>>

The Global Burden of Disease (GBD) 2019 study reported that microbial pathogens are responsible for more than 400 million years of life lost annually across the globe, a higher burden than either cancer or cardiovascular disease \citep{vos_global_2020}. In particular, lower respiratory infections, diarrhoeal diseases, HIV/AIDS and tuberculosis were amongst the five leading causes of global total years of life lost. More recently, the COVID-19 pandemic, declared as such by the World Health Organization (WHO) on 11 March 2020 after the emergence and global spread of the severe acute respiratory syndrome coronavirus 2 (SARS-CoV-2), and as of January 2022 has caused more than 5.63 million deaths worldwide \citep{ritchie_coronavirus_2020}, making it one of the deadliest pandemics in history. Coronavirus has been responsible for three of the eighteen major pandemics registered throughout modern history \citep{piret_pandemics_2021}, all occurring after the year 2000. \textit{Yersinia pestis}, responsible for three pandemics of plague, \textit{Vibrio cholerae}, with seven cholera pandemics, and Influenza A virus, the causative agent of five flu pandemics, are responsible for the remaining, with Influenza being the only other pathogen with a pandemic registered after 2000. Recent decades have also witnessed the emergence of additional virulent pathogens, including the Ebola virus, West Nile virus, Dengue virus and Zika virus, particularly in lower-income countries.

@@

The past few decades have seen a major revolution in the operation of microbial laboratories, driven by the development of molecular technologies and ways to make these accessible, namely amplification-based polymerase chain reaction (PCR), matrix-assisted laser desorption/ionisation - time of flight (MALDI-TOF) and DNA-microarray-based hybridisation technology. These are used in conjunction with traditional techniques such as microscopy, culture and serology, not fully replacing them. Application of these methods differs by suspected infection type: bacterial, viral, fungal or parasitic. For the purpose of this dissertation work, we will be focusing on bacterial and viral infections.

@@

The current gold standard for bacterial pathogen identification in diagnostic microbiology laboratories involves the isolation of the pathogen through culture followed by biochemical testing, a multi-step process that can take days to weeks before obtaining results, depending on the fastidiousness of the organism and if it can be cultured \citep{muhamad_rizal_advantages_2020, giuliano_guide_2019, muhamad_rizal_advantages_2020}. Although culturing allows the identification of a wide variety of organisms, some pathogens can escape routine investigation due to strict metabolic necessities for growth or the requirement for specific biochemical tests needed for their identification. Additionally, results will be obscured if a mixed culture is obtained, particularly if the cultures are obtained from sites with a microbiota, such as the gut and the skin, increasing the risk of contamination from normal flora, and leading to false results \citep{giuliano_guide_2019}. After successful growth in culture, Gram staining and MALDI-TOF mass spectrometry are often used for identification with good accuracy as long as the pathogen is presented in the the coexisting database \citep{patel_maldi-tof_2015}. An alternate rapid identification method is PCR where nucleic acid fragments are detected through specific probes, being highly sensitive and specific, to the point where PCR may detect bacteria that are not viable after a patient has been treated for an infection and it is limited to the probed used \citep{scerbo_beyond_2016}. Syndromic panels, an extension of PCR by using multiple primers (multiplex PCR) to simultaneously amplify the nucleic acids of multiple targets in a single reaction, tried to address this issue by allowing for the identification of multiple bacteria and other important information such as the detection of antibiotic resistance or virulence genes \citep{giuliano_guide_2019}

@@

Choosing an appropriate bacterial typing technique for epidemiological studies depends on the resources available and the minimum intended resolution, ranging from DNA fingerprinting to multilocus sequence typing, Pulsed-field gel electrophoresis (PFGE) and sequence-based typing (see \secref{sec:genomics_approach})  \citep{allerberger_molecular_2012,foxman_choosing_2005}. DNA macrorestriction analysis by PFGE, which revolutionised precise separation of DNA fragments, became the most widely implemented DNA fingerprinting technique \citep{allerberger_molecular_2012}, becoming the golden standard for bacterial typing \citep{neoh_pulsed-field_2019}.

In the early 2000s, Multilocus sequence typing (MLST) was proposed as a portable, universal, and definitive method for characterising bacteria \citep{maiden_multilocus_2006}. Instead of enzyme restriction of bacteria DNA, separation of the restricted DNA bands using a PFGE chamber, followed by clonal assignment of bacteria based on banding patterns, MLST relies on the amplification through PCR sequences of internal fragments of housekeeping genes (usually 5 to 7), approximately 450-500 basepairs (bp) in size, followed by its the sequence, usually my Sanger methods (see \secref{ssec:1st_gen_seq}). For each house-keeping gene, the different sequences present within a bacterial species are assigned as distinct alleles and, for each isolate, the alleles at each of the (usually) seven loci define the allelic profile or sequence type \citep{larsen_multilocus_2012}. As with PFGE, different schemes, defining what house-keeping gene fragments are used, are available depending on the species. Unlike PFGE, the provision of freely accessible, curated databases of MLST nucleotide sequence data enables the direct comparison of bacterial isolates, providing the basis of a common language for bacterial typing \citep{maiden_multilocus_2006}. So far, MLST schemes for 115 bacterial organisms have been published and made freely available\footnote{\url{https://pubmlst.org/organisms}}, \cite{jolley_open-access_2018})

@@

Direct detection methods negate the need for virus propagation, detecting the virus directly from the suspect source through nucleic acid and immunological methods. PCR and reverse transcription-PCR (RT-PCR) are widely applied methods for the detection of both DNA and RNA viruses, respectively, driven by increased awareness of the clinical value of, and demand for, prompt information about viral loads, viral sequence data, and potential antiviral resistance information \citep{cassedy_virus_2021}. Syndromic testing (see \secref{sssec:bacterial}) is now fully integrated into the standard testing practices of many clinical laboratories \citep{dien_bard_panels_2020}. Limitations of these assays include no detection of off-target pathogens, a lack of full susceptibility information, cost, and false-positive results. Real-time quantitative PCR (qPCR) remains the front line tool in aetiological diagnosis, measuring the production of the target amplicon throughout the reaction and providing quantitative results with high specificity and sensibility, albeit with a significant cost due to sophisticated apparatus despite high-throughput systems being widely established \citep{cassedy_virus_2021}.

Immunoassays employ singular-epitope specificity antibodies as the primary means to detect viruses within a sample and provide a much more cost-efficient alternative to nucleic acid detection \citep{cassedy_virus_2021}. One major application is seroprevalence assays, an essential technique for identifying patients who have been exposed to a virus (historical exposure), detecting asymptomatic infection or evaluating vaccine efficacy  \citep{chan_determining_2021, bobrovitz_global_2021}. Lateral flow immunoassays (LFIA) are extensively used for detecting virus-associated protein directly from the source through labelled antibodies binding to their cognate antigens, usually read by way of a colour change at a test line. Besides being very cost-effective, LFAIs have a turnaround time of minutes and the colour change can be observed with the naked eye, therefore facilitating rapid diagnosis but its results are limited to semi-quantitative and it does not typically achieve sensitivity comparable to nucleic-acid detection \citep{estrela_lateral_2016, cassedy_virus_2021, di_nardo_ten_2021}.

@@

Since the publication of the first complete microbial genome a quarter of a century ago, that of the bacterium \textit{Haemophilus influenzae} \citep{hood_dna_1996}, genomics has transformed the field of microbiology, and in particular its clinical application. The paper describing the DNA-sequencing method with chain-terminating inhibitors used in the sequencing of the first microbial genome \citep{sanger_dna_1977}, which earned the late Frederick Sanger his share of the 1980 Nobel Prize in Chemistry alongside Walter Gilbert, was, in 2014, the top fourth in the number of citations with 60335, highlighting its impact in the field of biological sciences, and by extension medicine \citep{van_noorden_top_2014}. Currently, this number has increased to 84546 according to PubMed Central\textsuperscript{\small\textregistered} (PMC)\footnote{\url{https://pubmed.ncbi.nlm.nih.gov/}}\footnote{\url{ https://www.ncbi.nlm.nih.gov/pmc/articles/PMC431765/}}. Since its emergence, reductions in cost, technical advances in sequencing technologies and new computational developments have made genomic sequencing one of the most influential tools in biomedical research, yielding unprecedented insights into microbial evolution and diversity, and the complexity of the genetic variation in both commensal and pathogenic microbes. The emerging application of genomic technologies in the clinic to combat infectious diseases is transforming clinical diagnostics and the detection and surveillance of outbreaks.

@@

The latter is defined by their use of terminator molecules that are similar to those used in the first-generation of sequencing, preventing elongation of the DNA molecule, but unlike the first methods, it is reversible. To begin the process, a DNA template is primed by a sequence that is complementary to an adapter region, which will initiate polymerase binding to this double-stranded DNA region. During each cycle, a mixture of all four individually labelled and 3'-blocked dNTPs are added. After the incorporation of a single dNTP to each elongating complementary strand, unbound dNTPs are removed and the surface is imaged to identify which dNTP was incorporated at each cluster by optical capture. The fluorophore and blocking group can then be removed and a new cycle can begin \citep{goodwin_coming_2016}. The Illumina systems, which use this technology, accounts for the largest market share for sequencing instruments compared to other platforms\footnote{\url{https://www.forbes.com/companies/illumina/?sh=774358a91aa6}}, allowing paired-end sequencing and having the highest throughput (from 25 million reads for a MiSeq instrument to  1.2 billion reads for a NextSeq instrument\footnote{\url{https://www.illumina.com/systems/sequencing-platforms.html}}), with read lengths ranging from 45 to 300 bases in length with high accuracy, albeit with long running times (4 to 55 hours), rendering this technology a good choice for many sequencing applications where large read length is not required \citep{loman_performance_2012, gupta_chapter_2014, hagemann_overview_2015}.

@@

Despite their wide adoption, second-generation methods require library preparation and an enrichment or amplification step. These steps are time-consuming, introduce biases related to preferential capture or amplification of certain regions, and produce reads with relatively small size, making transversing repetitive genomic regions impossible if they are larger than the read length \citep{hagemann_overview_2015}. Third-generation sequencing technologies, also known as long-read sequencing or single-molecule sequencing, are characterised by the generation of ultra-long-reads, albeit at a much lower throughput than the second-generation \citep{hoang_long-reads-based_2022}. They also have the potential to go beyond four-base sequencing to reveal genome-wide patterns of methylation and other chemical modifications that control the biology of bacteria or the virulence of pathogens \citep{korlach_going_2012}. Currently, commercial long-read sequencing is supported by two companies: Pacific Biosciences\footnote{\url{https://www.pacb.com/}} and Oxford Nanopore Technologies\footnote{\url{https://nanoporetech.com/}}.

@@

Whole-genome sequencing (WGS) is becoming one of the most widely used applications of microbial genome sequencing. The major advantage of WGS is to yield all the available DNA information content on isolates in a single rapid step following culture (sequencing without culture will be discussed in the \secref{ssec:metagenomics}). In principle, after obtaining a pure culture, either bacterial (see \secref{sssec:bacterial}) or viral (see \secref{sssec:viral}), the data from sequencing contain all the information currently used for diagnostic and typing needs, and much more, thus opening the prospect for large-scale research into pathogen genotype-phenotype associations from routinely collected data \citep{didelot_transforming_2012}.
The cost of producing massive amounts of information requires a new framework with expert handling and processing of computer-driven genomic information, as well as capable computational infrastructures, but through this technology, researchers and clinicians can obtain the most comprehensive view of genomic information and associated biological implications, transforming clinical diagnostics and the detection and surveillance of outbreaks. \citep{cirulli_uncovering_2010, nature_reviews_genetics_genomic_2019, goodwin_coming_2016}.

Targeted sequencing is also proving invaluable to clinical microbial and research, not only by allowing more individual samples to be sequenced within a single run, significantly reducing costs and the amount of data generated, but also, due to the smaller target size, obtaining results with very high confidence due to the high coverage obtained \citep{goodwin_coming_2016}. This has been particularly useful in viral genomics where sections, such as the capsid, or the complete viral genome can be selectively targeted directly from the suspected sample, offering a more time-effective method to achieve the same output as traditional nucleic acid amplification methods \citep{cassedy_virus_2021}.

@@

WGS has been used in the routine laboratory workflow when typing of pathogens by a method having the highest possible discriminatory power is required either through single nucleotide polymorphism (SNP) or core-genome/whole genome MLST (cg/wg MLST) analysis, for example during hospital outbreaks \citep{tagini_bacterial_2017}. Additionally, in bacterial diagnostics, WGS can be used to reveal the presence of AMR genes, or genes associated with virulence and pathogenicity, as well as to discover new genetic mechanisms for the three previously defined important clinical features of a bacterium \citep{rossen_practical_2018}. The implementation of WGS in routine diagnostics requires several adaptations in the laboratory workflow, from the ‘wet’ laboratory part (extraction, library preparation, sequencing), to the ‘dry’ bioinformatics part where genomic data is analysed and its results interpreted by specialised personnel \citep{rossen_practical_2018}.

@@

In outbreak detection and surveillance, genetic sequencing techniques combined with epidemiological data have undoubtedly provided immeasurable insights regarding evolutionary relationships and transmission pathways in various environments \citep{beckett_pandemic_2021, lancet_genomic_2021}. In a pandemic setting, this approach, although not novel, has been revolutionary, particularly in the COVID-19 setting. 
In the 2009 swine-origin Influenza A H1N1 pandemic, the first complete genome was publicly available on the 25 of April of 2009 (GenBank accession number FJ966079), about a month after records of increased flu activity in Mexico and 10 days after the first confirmed cases in California, United States of America \citep{smith_origins_2009, novel_swine-origin_influenza_a_h1n1_virus_investigation_team_emergence_2009}. By the time the pandemic was declared, on 11 of June of 2009, \cite{smith_origins_2009} reported the origins and evolutionary genomics of the pandemic influenza A variant with a collection of 813 complete influenza genome sets, 17 of which belonging to the newly swine influenza viruses (GenBank accessions numbers GQ229259–GQ229378). The MERS pandemic, declared as such in 2015 \citep{piret_pandemics_2021}, had its first publicly available sequence on 5 of July 2015 (GenBank accession number KT006149)\citep{lu_complete_2015}, with a sequence from a camel, thought to be an intermediate host for the virus, available as early as 7 of March 2016 (GenBank accession number KU740200) \citep{kandeil_complete_2016, al-shomrani_genomic_2020}.

@@

Despite the increasing adoption of DNA sequencing methods in clinical microbiology, the sequencing of genetic material from a pure culture requires \textit{a priori} knowledge of what to expect from a particular clinical sample or patient \citep{schuele_future_2021}. In most cases, this knowledge is enough to request the most appropriate test, such as multiplexed panels or specific culture media, but this is not always the case. In recent years, there has been a growing interest in using metagenomics to deliver culture-independent approaches to microbial ecology, surveillance and diagnosis \citep{loman_twenty_2015, loman_high-throughput_2012}.
Metagenomic DNA sequence allows detailed characterisation of pathogens in all kinds of samples originating from humans, animals, food and the environment, ligating the diagnostics to surveillance in a true "one health" fashion \citep{rossen__2018}. Unlike PCR or microarrays, it usually does not require primer or probe design, it can be easily multiplexed, and the specificity and selectivity of the sequencing can be adjusted computationally after acquiring the data \citep{dunne_next-generation_2012}.  While most molecular assays target only a limited number of pathogens, metagenomic approaches characterise all DNA or RNA present in a sample, enabling analysis of the entire microbiome as well as the human host genome or transcriptome in patient samples \citep{chiu_clinical_2019}. Whether or not it can entirely replace routine microbiology depends on several conditions and future developments, both technological and computational (see \secref{sec:bioinformatics}).

@@

Quality assessment and control is a basal step to any analysis, and aims to (1) remove and/or filter low quality and low complexity reads, (2) trim adapters, and (3) remove host sequences from the samples’ raw data. There are many tools available but the most commonly used are FastQC\footnote{\url{https://www.bioinformatics.babraham.ac.uk/projects/fastqc/}} (Babraham Bioinformatics) for quality control, followed by Trimmomatic \citep{bolger_trimmomatic_2014}, Cutadapt \citep{martin_cutadapt_2011} or fastp \citep{chen_fastp_2018} to trim and/or filter adaptors, low quality and low complexity sequences. For long-read sequencing, tools like NanoPlot and NanoStats \citep{de_coster_nanopack_2018}, and Filtlong\footnote{\url{https://github.com/rrwick/Filtlong/}} can perform the equivalent quality assessment and control, adapter trimming and low quality trimming, respectively.

@@

Metataxonomics (see \secref{sssec:metataxonomics}) is the most widely used technique for microbial diversity analysis \citep{hilton_metataxonomic_2016}, and due to its particularities, the analysis of this data is also very particular. Data analyses are mostly carried out through specialised pipelines that wrap and combine several tools, offering the possibility to follow a simple protocol with default configurations or choose between a plethora of different configurations to adjust for any particular needs. Quantitative Insights Into Microbial Ecology 2 (QUIIME2)\footnote{\url{https://qiime2.org/}} \citep{bolyen_reproducible_2019} has become the \textit{de facto} tool for metataxonomic analysis as a framework with an ever-growing suite of plugins and intuitive data visualisation tools for the assessment of results. Mothur \citep{schloss_introducing_2009} and UPARSE \citep{edgar_uparse_2013} are also a popular alternative although resulting outputs differing significantly between pipelines despite using the same inputs having been reported by \cite{marizzoni_comparison_2020}, with a magnitude that is comparable to differences in upstream sample treatment and sequencing procedures.

A typical workflow starts with quality filtering, error correction and removal of chimeric sequences. These quality control steps are followed by either taxonomic assignment of reads or a clustering step where reads are gathered into OTUs given their sequence identity, followed by statistical analysis to assess differences between given groups.

Taxonomic assignment methods classify query sequences based on the best hit found in reference databases of annotated sequences, being heavily dependent on the completeness of the reference databases (see \secref{sssec:metataxonomics}). Classification is further limited by lack of species annotation in most reference databases \citep{westcott_novo_2015}. Alternatively, the same approach of direct taxonomic classification, without OTU clustering, can be followed as with genomic and shotgun metagenomic data, given that the databases include rRNA sequences.

OTU clustering methods can be categorised into: (1) computationally expensive hierarchical methods that cluster sequences based on a distance matrix measuring the difference between each pair of sequences, (2) less expensive heuristic methods cluster sequences into OTUs based on a pre-defined threshold, generally, with a sequence being selected as a seed and the rest of the sequences being analysed sequentially and added to existing or new clusters according to the defined threshold, and (3) model based clustering methods that do not rely on a pre-defined and fixed threshold, defining OTUs based on a soft threshold and carrying out the clustering process based on methods such as an unsupervised probabilistic Bayesian clustering algorithm \citep{hao_clustering_2011}. These methods offer the possibility to cluster sequences based on criteria that do not depend on reference databases and are especially useful in less characterised microbial communities or with a high representation of uncultured microbes. Due to the assumptions made with this strategy, it is sensitive to under or overestimation of the number of OTUs in a sample as defining a threshold to accurately cluster sequences is difficult \citep{westcott_novo_2015}.
